---
title: "ENV 782 Final - CA Wildfires"
author: "Sky Volz and Brianna Karson"
date: "2026-04-24"
output: pdf_document
---

\tableofcontents
\listoffigures

# Rationale and Introductions

Wildfires present a pressing environmental challenge in California that has been exacerbated by decades of fire suppression, increasing drought conditions, and a changing climate. This report analyzes wildfire data from the past 75 years, focusing on how climate change has impacted wildfires.  Increasingly intense droughts may have a profound affect on Californian fire ecology, contributing to greater fire burn areas, and increased frequency of storms may be a driving higher frequencies of fires caused by lightning striking dry fuels. By providing statistical evidence that wildfire and lightning strike intensity has increased over time, this research can inform policy recommendations, ideally encouraging more proactive fire management strategies to mitigate threats to Californian landscapes. 

# Research Questions

This project explores the following research questions: 

  **1. How has the size of wildfire burn areas of changed over time?**
  
  **2. How has the number of wildfires caused by lightning strikes changed over time?**

# Data Information

This dataset is sourced from the California Natural Resources Agency GIS database and is serviced by the California Department of Forestry and Fire Protection’s  CAL FIRE eGIS. It includes layers that subset from the larger dataset according to year the fire occurred, and size of the burn area in acres. It contains 17,540 records, spanning from the 1950s to May 9, 2025. It is a publicly available feature layer and was last updated on May 9, 2025.

```{r setup, include=FALSE, message=FALSE, warning=FALSE}

#loading packages 
library(tidyverse); library(dplyr); library(here); library(lubridate)
library(sf); library(ggplot2); library(scales);
library(trend); library(zoo); library(Kendall); library(tseries); library(forecast)

#set working directory
here()
```

```{r data wrangling, message=FALSE, warning=FALSE, include=FALSE}
#reading in data
CA_fires <- read.csv(file=here(
  "Data/raw/California_Historic_Fire_Perimeters_1516624541847049096.csv"), 
  colClasses = c('Cause' = 'factor',
                 'Collection.Method' = 'factor',
                 'Management.Objective' = 'factor',
                 'State' = 'factor',
                 'Agency' = 'factor',
                 'Unit.ID' = 'factor'))

#formatting date columns
CA_fires$Alarm.Date <- mdy_hms(CA_fires$Alarm.Date)
CA_fires$Containment.Date <- mdy_hms(CA_fires$Containment.Date)

CA_fires <- CA_fires %>%
  mutate(Cause = recode(Cause, 
                        "1" = "Lightning",
                        "2" = "Equipment Use",
                        "3" = "Smoking",
                        "4" = "Campfire",
                        "5" = "Debris",
                        "6" = "Railroad",
                        "7" = "Arson",
                        "8" = "Playing With Fire",
                        "9" = "Miscellaneous",
                        "10" = "Vehicle",
                        "11" = "Powerline",
                        "12" = "Firefighter Training",
                        "13" = "Non-Firefighter Training",
                        "14" = "Unidentified",
                        "15" = "Structure",
                        "16" = "Aircraft",
                        "17" = "Volcanic",
                        "18" = "Escaped Prescribed Burn",
                        "19" = "Illegal Alien Campfire"
                        ))


#reading in shapefile
fires_shapefile <- st_read("Data/raw/California_Fire_Perimeters_(1950+).shp")
CA_boundary_shp <- st_read("Data/spatial/CA_State.shp")

write.csv(CA_fires, file=here("./Data/processed/CA_fires.csv"))
```

The following is a summary of column names and the first few rows of the dataset. 

```{r data summary, echo=FALSE}
#View dimensions, column names, variable type, and head of each column:
#str(CA_fires)

head(CA_fires)
#this prints really weird - maybe we just do the column names
```

\newpage

# Exploratory Data Analysis

Exploratory analyses were conducted in order to highlight possible trends or patterns in the data. The following plots reveal how California wildfire sizes and frequency have shifted over the past 75 years, along with leading fire causes. 

```{r theme set, include = FALSE}
mytheme <- theme_bw(base_size = 12) +
  theme(axis.text = element_text(color = "black"),legend.position = "top", 
        plot.title = element_text(face = "bold"))
theme_set(mytheme)
```

## Change Over Time
```{r histogram of fire counts, echo=FALSE, fig.cap= "Number of Fires Over Time", error=FALSE}
fire_count_plot <- CA_fires %>%
  ggplot(aes(x = Year)) + 
  labs( x = "Year",
        y = "Number of Fires",
        title = "Number of Fires Over Time") + 
  geom_histogram(binwidth = 1,
                 fill = 'firebrick3',
                 color = 'darkred') + 
  scale_x_continuous(n.breaks = 7)

print(fire_count_plot)
```

This chart plots the number of fires per year for the past 75 years, indicating a possible upward trend in the number of fires over time, with some notable spikes in years when the state exhibited unusually low or high fire activity. 

\newpage
```{r scatter plot fire burn size, error=FALSE, warning=FALSE, echo=FALSE, fig.cap= "Burn Size (Acres) by Year"}

options(scipen = 999) #this command keeps the y axis from switching into scientific notation

fire_size_plot <- CA_fires %>% 
  ggplot(aes(y=GIS.Calculated.Acres, x=Year)) + 
  geom_point(fill = 'firebrick3', color = 'darkred', alpha = 0.5) + 
  scale_y_continuous(labels = label_comma(),  limits = c(0,500000)) +
  labs(y="Burn Acreage (Calculated in GIS)", title = "Burn Size (Acres) by Year")

print(fire_size_plot)
```

This chart plots burn acreage over time, with the removal of a few outlier points. Though the majority of fires covers less than 50,000 acres, there appears to be an increase in points within the 300,000 to 500,000 burn acreage range since 1990. This is an indicator that California's wildfire burn acreage may be increasing over time.

\newpage
## Fire Causes
```{r fire causes scatter, echo=FALSE, fig.cap = "Fires Started by Each Cause, 1950 - 2025", error=FALSE, warning=FALSE}

fire_cause_scatter <- CA_fires %>% 
  ggplot(aes(y = Year, x = Cause)) + 
  geom_point(color = 'firebrick3', alpha = 0.1, size = 2.5) + 
  labs(y="Year", 
       title = "Fires Started by Each Cause, 1950 - 2025") +
  theme(axis.text.x = element_text(angle = 45, , vjust = 1, hjust = 1)) 
  
print(fire_cause_scatter)
```

This chart plots wildfire events by ignition cause over time, with each point representing a single fire from 1950–2025. Dense vertical groupings highlight causes that occur more frequently, while sparser columns represent less common ignition sources. Causes that span the full time range appear consistent across decades (Debris and Equipment Use), whereas those concentrated in certain periods (Vehicles and Powerlines) suggest more episodic occurrence or changing patterns over time.

\newpage
```{r bar graph of fire causes,echo=FALSE, fig.cap = "Total Fires Started by Each Cause", error=FALSE, warning=FALSE}

fire_cause_count <- CA_fires %>% 
  ggplot(aes(x = fct_infreq(Cause))) +
  geom_bar(fill = "firebrick3", color = "darkred") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Cause",
       y = "Number of Fires",
       title = "Total Fires Started by Each Cause")

print(fire_cause_count)
```

This plot reveals that the top five most commonly observed fire causes are: \newline
1. Unidentified \newline
2. Lightning \newline
3. Miscellaneous \newline
4. Equipment Use \newline
5. Arson \newline

Although the metadata does not describe what is meant by Unidentified or Miscellaneous, these categories are too large to ignore. These categories have likely persisted as a result of the vast variety of circumstances that may cause a fire to ignite, along with insufficient resources to conduct a robust investigation surrounding the cause of every small fire. 

\newpage
```{r box plots of fire causes, echo=FALSE, fig.cap = "Top Five Fire Causes Over Time", error=FALSE, warning=FALSE}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Unidentified" | Cause == "Lightning" | Cause == "Miscellaneous" | Cause == "Equipment Use" | Cause == "Arson") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year, color = Cause)) + 
  geom_line(linewidth = 0.75) +
  geom_point() +
  scale_color_brewer(palette = 'Paired') +
  labs(y="Number of Fires", 
       x="Year",
       title = "Top Five Fire Causes Over Time")  + 
  theme(legend.position =  "right")

print(fire_cause_plot)
```

This plot notably reveals that there appears to be an overall increasing trend in the fires caused by lightning, which is the second leading cause of fires in California. Trend analysis for lighting-caused fires is conducted later in this report. 

\newpage

## Mapping Burn Areas
```{r mapview, echo=FALSE, fig.cap="Map of California Wildfires, 1950 - 2025", error=FALSE, warning=FALSE}
# visualize overlapping fire areas using symbology to show where fires are concentrated

fire_map <- ggplot() + 
  geom_sf(data = CA_boundary_shp) +
  geom_sf(data = fires_shapefile, fill = 'firebrick3', color = 'darkred', alpha = 0.25) +
  labs(title = "Map of California Wildfires, 1950 - 2025")

print(fire_map)
```

This map provides an overview of all fire burn areas for the past 75 years. Areas in darker red indicate fire burn area overlap, thus highlighting regions that have been burned by multiple fires throughout the years. Burn areas are notably concentrated around mountain ranges, for the most part excluding California's central valley. This map is consistent with typical fire behavior, providing a visual depiction of how fire always moves uphill.

\newpage

# Analysis
These analyses evaluate how wildfire activity has changed over time, focusing on total burned area and fires caused by lightning. For burned area, fire burn acreage was added up to get monthly and yearly totals. The data was then analyzed at both scales to see if any change over time could be attributed to seasonal patterns. For the yearly data, a Mann-Kendall test was conducted in order to detect the presence of monotonic trends.

For the monthly burned area data, a time series decomposition was used to split the data into trend, seasonal patterns, and random variation. A Seasonal Mann–Kendall test was run in order to test for seasonal trends in the data.

The same approach was used for lightning-caused fires. Monthly and yearly fire counts were summed and analyzed over time. The yearly data was evaluated using a Mann–Kendall test, since there is no seasonal pattern at that scale. Monthly data was again broken down using decomposition, and a Seasonal Mann–Kendall test was used to detect trends while accounting for seasonal changes.

## Question 1: How has the size of wildfire burn areas of changed over time?
Null hypothesis: There has been no significant change in burn acreage of California wildfires over time. \newline

**Alternative Hypothesis: The burn acreage of California wildfires has significantly increased over time.** \newline

The first analysis focuses on whether fire acreage in California has increased, either as a result of an increase in overall fires or an increase in geographically larger fires. A time series was created for the yearly burn acreage over time and then plotted. Then, a Mann Kendall test and linear regression were run in order to identify any trend in the yearly data. Finally, fire acreage was calculated per month to create a time series analysis focused on seasonal trend analysis.

### Burn Area Yearly Time Series
```{r burn acreage yearly time series plot, echo=FALSE, fig.cap="Time Series Plot: Yearly Burn Acreage Over Time", warning=FALSE, message=FALSE}
#adding up acres burned per year and checking that every year is in there
fire_yearly <- CA_fires %>% 
  group_by(Year) %>% 
  summarise(total_acres = sum(GIS.Calculated.Acres, na.rm = TRUE)) %>% 
  complete(Year = 1950:2025, fill = list(total_acres = 0))

# creating annual time series
fire_acres_ts <- ts(fire_yearly$total_acres,
                    start = 1950,
                    end = 2025,
                    frequency = 1)

fire_yearly_ts_plot <- fire_yearly %>%
  ggplot(aes(x = Year , y = total_acres)) +
  geom_line(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick2') +
  labs(y = "Yearly Burn Acreage",
       title = "Time Series Plot: Yearly Burn Acreage Over Time")
print(fire_yearly_ts_plot)
``` 
\newpage

### Mann-Kendall
```{r burn acreage yearly time series mann kendall, echo=FALSE, error=FALSE, warning=FALSE}
#mann kendall
non_seasonal_acreage <- MannKendall(fire_acres_ts)
summary(non_seasonal_acreage)
```

### Linear Regression
```{r burn acreage yearly time series linear regression, echo=FALSE, error=FALSE, warning=FALSE}
# can't decompose yearly sum because its not seasonal so running a linear regression
# chose tslm over lm because tslm accounts for trends and lm does not
fire_lm <- tslm(fire_acres_ts ~ trend)
summary(fire_lm)
```
\newpage
### Burn Area Monthly Time Series
```{r burn acreage monthly time series, echo=FALSE, warning=FALSE, fig.cap="Time Series Plot: Monthly Burn Acreage Over Time", error=FALSE, message=FALSE}
#now looking at monthly trend over time - seasonal so I can decompose it
CA_fires_2 <- CA_fires %>% 
  mutate(Alarm.Date = as.Date(Alarm.Date)) %>%
  mutate(Month = floor_date(Alarm.Date, unit = "month")) %>% #creates a monthly grouping
  group_by(Month) 

#creatly monthly sum
fire_monthly <- CA_fires_2 %>%
  summarise(total_acres = sum(GIS.Calculated.Acres, na.rm = TRUE)) %>%
  ungroup()

#create time series
monthly_acres_ts <- ts(fire_monthly$total_acres, start = c(1950, 1), frequency = 12)

fire_monthly_ts_plot <- fire_monthly %>%
  ggplot(aes(x = Month , y = total_acres)) +
  geom_line(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick3') +
  labs(y = "Monthly Burn Acreage",
       title = "Time Series Plot: Monthly Burn Acreage Over Time")
print(fire_monthly_ts_plot)
```

\newpage
### Decomposition
```{r burn acreage monthly decomp, echo=FALSE, fig.cap= "Time Series Decomposition: Monthly Burn Acreage Over Time", warning=FALSE,  error=FALSE}
#decompose
monthly_acres_decomp <- stl(monthly_acres_ts, s.window = "periodic")
plot(monthly_acres_decomp,
      main = "Time Series Decomposition: Monthly Burn Acreage Over Time")
```

\newpage
### Seasonal Mann-Kendall
```{r burn acreage monthly smk, echo=FALSE, fig.cap= "Time Series Decomposition: Monthly Burn Acreage Over Time", warning=FALSE,  error=FALSE}
#seasonal = running seasonal mann-kendall test for trends
monthly_trend <- trend::smk.test(monthly_acres_ts)
monthly_trend
summary(monthly_trend)
```

### Summary
The Seasonal Mann–Kendall test of monthly fire burn acreage demonstrates variability and occasional increases in wildfire activity, but no significant trend in fire burn area over time at the monthly scale once seasonality is accounted for. The monthly time series decomposition supports this: the trend component shows a gradual upward slope with occasional sharp jumps that reflect extreme fire years and major shifts in conditions, but the seasonal component shows a stable, repeating annual fire cycle, and the remainder highlights large spikes in activity that represent unusually severe fire events. This contrasts with the yearly fire acreage trend discussed below.

The yearly fire burn area Mann-Kendall test and linear regression show a clear, statistically significant increase in burn acreage over time, with a positive coefficient denoting the increasing trend. The R² value of 0.1803 means that about 18% of the change in burn area sizes over time can be explained by time alone (p < 0.05). These results confirm the alternative hypothesis that the burn acreage of California wildfires has significantly increased over time.  Overall, the combined monthly and yearly results suggest that wildfire sizes are increasing over time, but much of the change is driven by long-term shifts and extreme fire events rather than a steady month-to-month increase. 

\newpage
## Question 2: How has the number of wildfires caused by lightning strikes changed over time?
Null hypothesis: There has been no significant change in the number of California wildfires caused by lightning over time. \newline

**Alternative Hypothesis: The number of California wildfires caused by lightning strikes has significantly increased over time** \newline

Informed by the exploratory analysis, these time series investigate whether there is sufficient statistical evidence to determine if the number of fires caused by lightning strikes has increased over time. A time series was created for the number of lightning-caused fires each year and then plotted. Then, a Mann Kendall test was run in order to identify any trend in the yearly data. Finally, the number of lightning-caused fires per month was calculated in order to create a time series analysis that breaks down seasonal trends.

\newpage
### Lightning-Caused Fires Yearly Time Series
```{r fires caused by lightning, echo=FALSE, fig.cap = "Number of Fires Caused by Lightning Strikes Over Time", error=FALSE, warning=FALSE, message=FALSE}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year)) + 
  geom_line(linewidth = 0.75, color = 'darkred') +
  geom_point(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick2') +
  labs(y="Number of Fires", 
       x="Year",
       title = "Number of Fires Caused by Lightning Strikes Over Time")  + 
  theme(legend.position =  "right")

print(fire_cause_plot)
```


### Mann-Kendall
```{r lightning yearly time series, echo=FALSE, warning=FALSE,  error=FALSE, message=FALSE}

lightning_fire_yearly <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  mutate(Year = as.factor(Year)) %>%
  group_by(Year) %>% 
  mutate(Year = as.Date(Year, format = "%Y"))  %>%
  summarise(count = n()) 

#creating annual time series
lightning_fire_yearly_ts <- ts(lightning_fire_yearly,
                    start = 1950,
                    end = 2025,
                    frequency = 1)

#removing time series plot because it matches the one from our exploratory analysis

#mann kendall
non_seasonal_lightning <- MannKendall(lightning_fire_yearly_ts)
summary(non_seasonal_lightning)
```

\newpage
### Lightning-Caused Fires Monthly Time Series
```{r lightning monthly time series2, echo=FALSE, fig.cap = "Time Series Plot: Monthly Fires Caused by Lightning Over Time", warning=FALSE, message=FALSE}
#now looking at monthly trend over time - seasonal so I can decompose it
lightning_fire_monthly <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  mutate(Alarm.Date = as.Date(Alarm.Date)) %>%
  mutate(Month = floor_date(Alarm.Date, unit = "month")) %>% #creates a monthly grouping
  group_by(Month) 

#create monthly sum
lightning_fire_monthly <- lightning_fire_monthly %>%
  summarise(count = n())

#creating data frame
Month <- as.data.frame(seq(from = as.Date("1950-05-01"),
                           to = as.Date("2025-01-01"), by = "month"))

#renaming column
Month <- rename(Month, 'Month' = 'seq(from = as.Date(\"1950-05-01\"), to = as.Date(\"2025-01-01\"), by = \"month\")')

#joining
lightning_fire_monthly_join <- left_join(Month, lightning_fire_monthly, by = "Month")

#linear interpolation
lightning_fire_monthly_join$count <- na.approx(lightning_fire_monthly_join$count, rule = 2)

#create time series
lightning_monthly_ts <- ts(lightning_fire_monthly_join$count, start = c(1950, 1), frequency = 12)

#plot time series
lightning_monthly_ts_plot <- lightning_fire_monthly_join %>%
  ggplot(aes(x = Month , y = count)) +
  geom_line(color = 'darkred') +
  labs(y = "Monthly Fire Count",
       title = "Time Series Plot: Monthly Fires Caused by Lightning Over Time")
print(lightning_monthly_ts_plot)
```

\newpage
### Decomposition
```{r lightning monthly decomp, echo=FALSE, fig.cap ="Time Series Decomposition: Monthly Count of Fires Caused by Lightning", warning=FALSE, }
#decompose
lightning_monthly_decomp <- stl(lightning_monthly_ts, s.window = "periodic")
plot(lightning_monthly_decomp,
     main = "Time Series Decomposition: Monthly Count of Fires Caused by Lightning")
```

\newpage
### Seasonal Mann-Kendall
```{r lightning monthly smk, echo=FALSE, fig.cap ="Time Series Decomposition: Monthly Count of Fires Caused by Lightning", warning=FALSE, }
#seasonal = running seasonal mann-kendall test for trends
lightning_monthly_trend <- trend::smk.test(lightning_monthly_ts)
lightning_monthly_trend
summary(lightning_monthly_trend)
```

### Summary
The Mann-Kendall trend test found that there has been a significant increase in the annual number of lightning-caused fires over time (tau = 0.111, p = 0.043). Moreover, the Seasonal Mann-Kendall test reveals that the monthly number of lightning-caused fires has also increased over time, even when normal seasonal patterns are taken into account (z = 7.42, p < 0.001). The increase was strongest during late winter and spring, especially in Seasons 2 through 5 (February - May), with Season 3 (March) showing the largest rise over time. 

The decomposition of the monthly time series shows that lightning-caused fires follow a regular yearly pattern, with more fires in some months and fewer in others. This means the timing of these fires is fairly consistent from year to year, following a seasonal pattern. It also shows that the overall number of lightning-caused fires has gradually increased over time, with more recent years generally having higher counts. The ups and downs of the remainder plot represent unusual months with especially high or low fire activity that are not consistent with the typical seasonal pattern.

Overall, these results demonstrate that the number of California wildfires caused by lightning strikes has significantly increased over time at both a monthly and yearly scale, thus confirming the alternative hypothesis.

# Conclusions
what this means for fires in CA and our focus on climate change 
maybe a policy recomendation or something??
This can help determine if fire trends are related to increasing lightning storms as a result of climate change.

# Limitations
It is noted in the metadata for this dataset that there are a number of fires missing from these records, mostly during the first few decades of data collection. As a result, certain trend analyses may present differently than if fire from the lats 75 years was able to be included. This is important to note when drawing conclusions from the data, particularly when focusing the on yearly and monthly fire acreage. 

The two categories of "Unidentified" and "Miscellaneous" also present a degree of limitation due to a lack of specificity. It would be interesting to run the exploratory analyses again, if the fires in these categories were able to be grouped correctly by their specific causes. 

## References
Fire Data: \newline
URL: https://gis.data.cnra.ca.gov/datasets/CALFIRE-Forestry::california-fire-perimeters-1950/about \newline 
Citation Title: Fire Perimeters (firep24_1) \newline 
Revision date: 2025-04-23 \newline 
Edition: 24_1 \newline 
Processing environment Version 6.2 (Build 9200); Esri ArcGIS 10.8.0.12790 

CA Boundary Shapefile:\newline 
URL: https://data.ca.gov/dataset/e212e397-1277-4df3-8c22-40721b095f33/resource/3db1e426-fb51-44f5-82d5-a54d7c6e188b/download/ca_state.zip \newline 
California State boundary in shapefile format from the US Census Bureau's 2023 MAF/TIGER database.