---
title: "ENV 782 Final - CA Wildfires"
author: "Sky Volz and Brianna Karson"
date: "2026-04-02"
output: pdf_document
---

\tableofcontents
\listoffigures

##Rationale and Introductions

Wildfires are a pressing environmental challenge in California that has been exacerbated by decades of fire suppression, increasing drought conditions, and a changing climate. This dataset contains wildfire data for the past 75 years, offering a unique glimpse into how fire patterns have shifted over time. This report focuses on how climate change has impacted wildfires, with increasingly intense droughts contributing to larger fires, or with increased frequency of storms potentially causing more fires via lightning striking dry fire fuels. By providing statistical evidence that wildfire or lightning strike intensity has increased over time, this research can inform policy recommendations, ideally encouraging more proactive fire management strategies such as prescribed burning to reduce fuels.

##Research Questions
1. Has the size of California wildfire burn areas of changed over time?
2. Has the frequency of California wildfires caused by lightning strikes changed over time?

##Dataset Information
This dataset is sourced from the California Natural Resources Agency GIS database and is serviced by the California Department of Forestry and Fire Protection’s  CAL FIRE eGIS. It includes layers that subset from the larger dataset according to year the fire occurred, and size of the burn area in acres. It contains 17,540 records, spanning from the 1950s to May 9, 2025. It is a publicly available feature layer and was last updated on May 9, 2025.

```{r setup, include=FALSE}
#loading packages 
library(tidyverse)
library(dplyr)
library(here)

#set working directory
here()
```

```{r data wrangling, include=FALSE}
#reading in data
CA_fires <- read.csv(file=here(
  "Data/raw/California_Historic_Fire_Perimeters_1516624541847049096.csv"), 
  colClasses = c('Cause' = 'factor',
                 'Collection.Method' = 'factor',
                 'Management.Objective' = 'factor',
                 'State' = 'factor',
                 'Agency' = 'factor',
                 'Unit.ID' = 'factor'))

#formatting date columns
CA_fires$Alarm.Date <- mdy_hms(CA_fires$Alarm.Date)
CA_fires$Containment.Date <- mdy_hms(CA_fires$Containment.Date)

CA_fires <- CA_fires %>%
  mutate(Cause = recode(Cause, 
                        "1" = "Lightning",
                        "2" = "Equipment Use",
                        "3" = "Smoking",
                        "4" = "Campfire",
                        "5" = "Debris",
                        "6" = "Railroad",
                        "7" = "Arson",
                        "8" = "Playing With Fire",
                        "9" = "Miscellaneous",
                        "10" = "Vehicle",
                        "11" = "Powerline",
                        "12" = "Firefighter Training",
                        "13" = "Non-Firefighter Training",
                        "14" = "Unidentified",
                        "15" = "Structure",
                        "16" = "Aircraft",
                        "17" = "Volcanic",
                        "18" = "Escaped Prescribed Burn",
                        "19" = "Illegal Alien Campfire"
                        ))
```

##Exploratory Data Analysis
```{r theme set}
#I'm not sure if you have preferences for theme, but we should set it here so that all the plots match

mytheme <- theme_minimal(base_size = 12) +
  theme(axis.text = element_text(color = "black"),legend.position = "top", 
        plot.title = element_text(face = "bold"))
theme_set(mytheme)
```

```{r histogram of fire counts}
#I was just messing around, let me know what you think of this! And the colors!
fire_count_plot <- CA_fires %>%
  ggplot(aes(x = Year)) + 
  labs( x = "Year",
        y = "Number of Fires",
        title = "Changes in Fire Occurence Over Time") + 
  geom_histogram(binwidth = 1,
                 fill = 'firebrick3',
                 color = 'darkred') + 
  scale_x_continuous(n.breaks = 7)

print(fire_count_plot)
```

```{r scatter plot of fire burn size}
fire_size_plot <- CA_fires %>% 
  ggplot(aes(y=GIS.Calculated.Acres, x=Year)) + 
  geom_point(fill = 'firebrick3', color = 'darkred', alpha = 0.5) + 
  #idk was just trying stuff with the colors
  labs(y="Burn Acreage Calculated in GIS") + 
  ylim(0,500000)  # I set limits to remove outliers, but do we want the outliers there? 

#also tried to add a trend line but because of how many small fires there are it was just flat parallel to the x axis

print(fire_size_plot)
```

```{r box plots of fire causes}
fire_cause_plot <- CA_fires %>% 
  ggplot(aes(y=Cause, x=Year, color = 'darkred')) + 
  geom_boxplot() + 
  labs(y="Cause of Fire", 
       x="Year",
       title = "Changes in Frequency of Fire Causes Over Time") + 
  theme(legend.position = "none")
print(fire_cause_plot)

#I tried to get the colors to match the histogram and it wont work and I don't know why

#should we exclude miscellaneous, illegal alien campfire (what?!), and unidentified ? 

#or maybe we can only show the the top 5 (for example, non-firefighter training and firefighter training seem sort of random and unnecessary) #or we can exclude those too? I feel like anything is fine as long we we justify why we did it
```

```{r mapview}
# visualize overlapping fire areas using symbology to show where fires are concentrated
```

##Hypotheses
We will write our hypotheses here

##Analysis
We will state our methods here
```{r burn acreage time series}

#Time Series Analysis: How has the burn acreage of wildfires changed over time?
  #Area as the dependent variable
  #Decomposition and trend analysis
```

```{r lightning time series}

#Time Series Analysis: Has the frequency or intensity of fires caused by lightning strikes changed over time?
  #Dependent variable is the count of lightning strikes
  #Decomposition and trend analysis
```

##Summary and Conclusions
#Results
#Discussion
#Limitations
see “Use Limitation” section in metadata

##References
Remember to cite dataset here 