---
title: "ENV 782 Final - CA Wildfires"
author: "Sky Volz and Brianna Karson"
date: "2026-04-19"
output: pdf_document
---

\tableofcontents
\listoffigures

# Rationale and Introductions

Wildfires present a pressing environmental challenge in California that has been exacerbated by decades of fire suppression, increasing drought conditions, and a changing climate. This dataset contains wildfire data from the past 75 years, offering a unique glimpse into how fire patterns have shifted over time. This report focuses on how climate change has impacted wildfires, with increasingly intense droughts contributing to larger fires, or with increased frequency of storms causing more and more fires due to lightning striking dry fuels. By providing statistical evidence that wildfire and lightning strike intensity has increased over time, this research can inform policy recommendations, ideally encouraging more proactive fire management strategies to mitigate threats to Californian landscapes. 

# Research Questions

This project explores the following research questions: 

  **1. Has the size of California wildfire burn areas of changed over time?**
  
  **2. Has the frequency of California wildfires caused by lightning strikes changed over time?**

# Dataset Information

This dataset is sourced from the California Natural Resources Agency GIS database and is serviced by the California Department of Forestry and Fire Protection’s  CAL FIRE eGIS. It includes layers that subset from the larger dataset according to year the fire occurred, and size of the burn area in acres. It contains 17,540 records, spanning from the 1950s to May 9, 2025. It is a publicly available feature layer and was last updated on May 9, 2025.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

#loading packages 
library(tidyverse); library(dplyr); library(here); library(lubridate)
library(sf); library(ggplot2); library(scales);
library(trend); library(zoo); library(Kendall); library(tseries); library(forecast)

#set working directory
here()
```

```{r data wrangling, message=FALSE, warning=FALSE, include=FALSE}
#reading in data
CA_fires <- read.csv(file=here(
  "Data/raw/California_Historic_Fire_Perimeters_1516624541847049096.csv"), 
  colClasses = c('Cause' = 'factor',
                 'Collection.Method' = 'factor',
                 'Management.Objective' = 'factor',
                 'State' = 'factor',
                 'Agency' = 'factor',
                 'Unit.ID' = 'factor'))

#formatting date columns
CA_fires$Alarm.Date <- mdy_hms(CA_fires$Alarm.Date)
CA_fires$Containment.Date <- mdy_hms(CA_fires$Containment.Date)

CA_fires <- CA_fires %>%
  mutate(Cause = recode(Cause, 
                        "1" = "Lightning",
                        "2" = "Equipment Use",
                        "3" = "Smoking",
                        "4" = "Campfire",
                        "5" = "Debris",
                        "6" = "Railroad",
                        "7" = "Arson",
                        "8" = "Playing With Fire",
                        "9" = "Miscellaneous",
                        "10" = "Vehicle",
                        "11" = "Powerline",
                        "12" = "Firefighter Training",
                        "13" = "Non-Firefighter Training",
                        "14" = "Unidentified",
                        "15" = "Structure",
                        "16" = "Aircraft",
                        "17" = "Volcanic",
                        "18" = "Escaped Prescribed Burn",
                        "19" = "Illegal Alien Campfire"
                        ))


#reading in shapefile
#fires_shapefile <- st_read("Data/raw/California_Fire_Perimeters_(1950+).shp")
#submit link here

#save this processed data table to processed folder - idk why this wont work
write.csv(CA_fires, file=here("./Data/processed/CA_fires.csv"))
```

# Exploratory Data Analysis

Our first step was to look at the data to look for any overall trends or notable patterns, to better inform our analysis. The following plots reveal how the fire sizes and frequency have occurred throughout the past 75 years, as well as what the top five fire causes are. 

```{r theme set, include = FALSE}
mytheme <- theme_bw(base_size = 12) +
  theme(axis.text = element_text(color = "black"),legend.position = "top", 
        plot.title = element_text(face = "bold"))
theme_set(mytheme)

```

```{r data summary, include = TRUE}
#View dimensions, column names, variable type, and head of each column:
str(CA_fires)

head(CA_fires)
```

```{r histogram of fire counts, fig.cap = "Number of Fires Over Time"}
fire_count_plot <- CA_fires %>%
  ggplot(aes(x = Year)) + 
  labs( x = "Year",
        y = "Number of Fires",
        title = "Number of Fires Over Time") + 
  geom_histogram(binwidth = 1,
                 fill = 'firebrick3',
                 color = 'darkred') + 
  scale_x_continuous(n.breaks = 7)

print(fire_count_plot)
```

notes on changes in frequency

```{r scatter plot of fire burn size, fig.cap = "Burn Size (Acres) by Year"}

options(scipen = 999) #this command keeps the y axis from switching into scientific notation

fire_size_plot <- CA_fires %>% 
  ggplot(aes(y=GIS.Calculated.Acres, x=Year)) + 
  geom_point(fill = 'firebrick3', color = 'darkred', alpha = 0.5) + 
  scale_y_continuous(labels = label_comma(),  limits = c(0,500000)) +
  labs(y="Burn Acreage (Calculated in GIS)", title = "Burn Size (Acres) by Year")
print(fire_size_plot)
```

notes on fire size plot - note concentration of small fires on bottom, interest in larger burn areas right side

```{r, fig.cap = "Total Fires Started by Each Cause, 1950 - 2025"}

fire_cause_scatter <- CA_fires %>% 
  ggplot(aes(y = Year, x = Cause)) + 
  geom_point(color = 'firebrick3', alpha = 0.1, size = 2.5) + 
  labs(y="Year", 
       title = "Total Fires Started by Each Cause, 1950 - 2025") +
  theme(axis.text.x = element_text(angle = 45, , vjust = 1, hjust = 1)) 
  
print(fire_cause_scatter)
```

This scatterplot is helpful because it shows wildfire events by ignition cause over time, with each point representing a single fire from 1950–2025. Denser vertical groupings indicate causes that occur more frequently, while sparser columns represent less common ignition sources. Causes that span the full time range appear persistent across decades (Debris and Equipment Use), whereas those concentrated in certain periods (Vehicles and Powerlines) suggest more episodic occurrence or changing patterns over time.

```{r bar graph of fire causes, fig.cap = "Total Fires Started by Each Cause, 1950 - 2025"}

fire_cause_count <- CA_fires %>% 
  ggplot(aes(x = fct_infreq(Cause))) +
  geom_bar(fill = "firebrick3", color = "darkred") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Cause",
       y = "Number of Fires",
       title = "Total Fires Started by Each Cause, 1950 - 2025")

print(fire_cause_count)
```

This plot reveals that the top five most commonly observed fire causes are:
  1. Unidentified 
  2. Lightning
  3. Miscellaneous
  4. Equipment Use
  5. Arson
  
Notes on what the unidentified and miscellanous sections are why they might be in the top 5

```{r box plots of fire causes, fig.cap = "Top Five Fire Causes Over Time"}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Unidentified" | Cause == "Lightning" | Cause == "Miscellaneous" | Cause == "Equipment Use" | Cause == "Arson") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year, color = Cause)) + 
  geom_line(linewidth = 0.75) +
  geom_point() +
  scale_color_brewer(palette = 'Paired') +
  labs(y="Number of Fires", 
       x="Year",
       title = "Top Five Fire Causes Over Time")  + 
  theme(legend.position =  "right")
print(fire_cause_plot)
```

This plot notably reveals that there appears to be an overall increasing trend in the fires caused by lightning, which is the second leading cause of fires in California. In the following plot, we focused in on the trend in lighting-caused fires, and there is a clear increase. This informs our focus on fires caused by lightning for the remainder of the analysis.

```{r fires caused by lightning, fig.cap = "Number of Fires Caused by Lightning Strikes Over Time"}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year)) + 
  geom_line(linewidth = 0.75, color = 'darkred') +
  geom_point(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick2') +
  labs(y="Number of Fires", 
       x="Year",
       title = "Number of Fires Caused by Lightning Strikes Over Time")  + 
  theme(legend.position =  "right")
print(fire_cause_plot)
```

```{r mapview, fig.cap="Map of Fires 1950 - 2025"}
# visualize overlapping fire areas using symbology to show where fires are concentrated
fire_map <- ggplot(data = fires_shapefile) + 
  geom_sf( fill = 'firebrick3', color = 'darkred', alpha = 0.5)

print(fire_map)
```

# Hypotheses
From this exploratory analysis, we developed the following hypotheses:
  **Alternative Hypothesis: The burn acreage of California wildfires has increased over time.**
    Null hypothesis: There has been no change in burn acreage of California wildfires over time.

  **Alternative Hypothesis: The number of California wildfires caused by lightning strikes has increased over time**
    Null hypothesis: There has been no change in the frequency of California wildfires caused by lightning over time.

# Analysis
We will state our methods here - note about why tslm was chosen over lm

Time Series Analysis: How has the burn acreage of wildfires changed over time?
  Area as the dependent variable
  Decomposition and trend analysis
  **first looking at yearly, then monthly, to see if the change / increase is seasonal or overall**

## Question One - Burn Acreage Over Time

```{r burn acreage yearly time series plot, fig.cap="Yearly Burn Acreage Over Time"}
#adding up acres burned per year and checking that every year is in there
fire_yearly <- CA_fires %>% 
  group_by(Year) %>% 
  summarise(total_acres = sum(GIS.Calculated.Acres, na.rm = TRUE)) %>% 
  complete(Year = 1950:2025, fill = list(total_acres = 0))

# creating annual time series
fire_acres_ts <- ts(fire_yearly$total_acres,
                    start = 1950,
                    end = 2025,
                    frequency = 1)

fire_yearly_ts_plot <- fire_yearly %>%
  ggplot(aes(x = Year , y = total_acres)) +
  geom_line(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick2') +
  labs(y = "Yearly Burn Acreage",
       title = "Yearly Burn Acreage Over Time")
print(fire_yearly_ts_plot)
``` 

```{r burn acreage yearly time series linear regression}
# can't decompose yearly sum because its not seasonal so running a linear regression
# chose tslm over lm because tslm accounts for trends and lm does not
fire_lm <- tslm(fire_acres_ts ~ trend)
summary(fire_lm)

#mann kendall
non_seasonal_acreage <- MannKendall(fire_acres_ts)
summary(non_seasonal_acreage)
```

```{r burn acreage monthly time series, fig.cap="Time Series Plot: Monthly Burn Acreage Over Time"}
#now looking at monthly trend over time - seasonal so I can decompose it
CA_fires_2 <- CA_fires %>% 
  mutate(Alarm.Date = as.Date(Alarm.Date)) %>%
  mutate(Month = floor_date(Alarm.Date, unit = "month")) %>% #creates a monthly grouping
  group_by(Month) 

#creatly monthly sum
fire_monthly <- CA_fires_2 %>%
  summarise(total_acres = sum(GIS.Calculated.Acres, na.rm = TRUE)) %>%ungroup()

#create time series
monthly_acres_ts <- ts(fire_monthly$total_acres, start = c(1950, 1), frequency = 12)

fire_monthly_ts_plot <- fire_monthly %>%
  ggplot(aes(x = Month , y = total_acres)) +
  geom_line(color = 'darkred') +
  labs(y = "Monthly Burn Acreage",
       title = "Time Series Plot: Monthly Burn Acreage Over Time")
print(fire_monthly_ts_plot)
```

```{r burn acreage monthly time series}
#decompose
monthly_acres_decomp <- stl(monthly_acres_ts, s.window = "periodic")
plot(monthly_acres_decomp,
      main = "Time Series Decomposition: Monthly Burn Acreage Over Time")

#seasonal = running seasonal mann-kendall test for trends
monthly_trend <- trend::smk.test(monthly_acres_ts)
monthly_trend
summary(monthly_trend)
```

from yearly trend linear regression:
  positive coefficent = increasing burned area over time
  r2 = 0.1803 is explained in change in fire acres explained by time alone
  p much <0.05 so statistically significant
  also shown in graphing

from monthly trend decomp:
  remainder: large spikes = extreme fire events
  trend: upward slope -> increasing fire severity over decades, sharp jumps -> major fire years / climate     shifts / policy changes?
  seasonal: Stable pattern = consistent fire season year-to-year

This Seasonal Mann–Kendall test shows weak evidence of an increasing trend in monthly wildfire acreage over time (z = 1.648), but the trend is not statistically significant (p = 0.099). Seasonal decomposition doesn't show a strong trend within individual months, but some of the summer months have slightly positive but non-significant increases. For example, July (tau = 0.148). These results show variability and occasional increases in wildfire activity, but not a large, consistent seasonal trend across all months. This contrasts with the trend shown when looking at yearly fire acreage.

## Question Two - Lightning Frequency Over Time
```{r lightning yearly time series, fig.cap = "Yearly Number of Fires Caused by Lightning"}

lightning_fire_yearly <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  mutate(Year = as.factor(Year)) %>%
  group_by(Year) %>% 
  mutate(Year = as.Date(Year, format = "%Y"))  %>%
  summarise(count = n()) 

#creating annual time series
lightning_fire_yearly_ts <- ts(lightning_fire_yearly,
                    start = 1950,
                    end = 2025,
                    frequency = 1)

#removing time series plot because it matches the one from our exploratory analysis

#mann kendall
non_seasonal_lightning <- MannKendall(lightning_fire_yearly_ts)
summary(non_seasonal_lightning)
```

```{r lightning monthly time series}
#now looking at monthly trend over time - seasonal so I can decompose it
lightning_fire_monthly <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  mutate(Alarm.Date = as.Date(Alarm.Date)) %>%
  mutate(Month = floor_date(Alarm.Date, unit = "month")) %>% #creates a monthly grouping
  group_by(Month) 

#create monthly sum
lightning_fire_monthly <- lightning_fire_monthly %>%
  summarise(count = n())

#creating data frame
Month <- as.data.frame(seq(from = as.Date("1950-05-01"),
                           to = as.Date("2025-01-01"), by = "month"))

#renaming column
Month <- rename(Month, 'Month' = 'seq(from = as.Date(\"1950-05-01\"), to = as.Date(\"2025-01-01\"), by = \"month\")')

#joining
lightning_fire_monthly_join <- left_join(Month, lightning_fire_monthly, by = "Month")

#linear interpolation
lightning_fire_monthly_join$count <- na.approx(lightning_fire_monthly_join$count)

#create time series
lightning_monthly_ts <- ts(lightning_fire_monthly_join$count, start = c(1950, 1), frequency = 12)
plot(lightning_monthly_ts)

#decompose
lightning_monthly_decomp <- stl(lightning_monthly_ts, s.window = "periodic")
plot(lightning_monthly_decomp)


#seasonal = running seasonal mann-kendall test for trends
lightning_monthly_trend <- trend::smk.test(lightning_monthly_ts)
lightning_monthly_trend
summary(lightning_monthly_trend)
```

# Results
should this be sprinkled throughout, then restated here?
# Summary and Conclusions
# Discussion
# Limitations
see “Use Limitation” section in metadata 
discuss the categories of miscellaneous and unidentified

## References
Remember to cite dataset here 