---
title: "ENV 782 Final - CA Wildfires"
author: "Sky Volz and Brianna Karson"
date: "2026-04-19"
output: pdf_document
---

\tableofcontents
\listoffigures

# Rationale and Introductions

Wildfires present a pressing environmental challenge in California that has been exacerbated by decades of fire suppression, increasing drought conditions, and a changing climate. This dataset contains wildfire data from the past 75 years, offering a unique glimpse into how fire patterns have shifted over time. This report focuses on how climate change has impacted wildfires, with increasingly intense droughts contributing to larger fires, or with increased frequency of storms causing more and more fires due to lightning striking dry fuels. By providing statistical evidence that wildfire and lightning strike intensity has increased over time, this research can inform policy recommendations, ideally encouraging more proactive fire management strategies to mitigate threats to Californian landscapes. 

# Research Questions

This project explores the following research questions: 

  **1. Has the size of California wildfire burn areas of changed over time?**
  
  **2. Has the frequency of California wildfires caused by lightning strikes changed over time?**

# Dataset Information

This dataset is sourced from the California Natural Resources Agency GIS database and is serviced by the California Department of Forestry and Fire Protection’s  CAL FIRE eGIS. It includes layers that subset from the larger dataset according to year the fire occurred, and size of the burn area in acres. It contains 17,540 records, spanning from the 1950s to May 9, 2025. It is a publicly available feature layer and was last updated on May 9, 2025.

The dataset can be found at this link: https://gis.data.cnra.ca.gov/datasets/CALFIRE-Forestry::california-fire-perimeters-1950/about  

```{r setup, include=FALSE, message=FALSE, warning=FALSE}

#loading packages 
library(tidyverse); library(dplyr); library(here); library(lubridate)
library(sf); library(ggplot2); library(scales);
library(trend); library(zoo); library(Kendall); library(tseries); library(forecast)

#set working directory
here()
```

```{r data wrangling, message=FALSE, warning=FALSE, include=FALSE}
#reading in data
CA_fires <- read.csv(file=here(
  "Data/raw/California_Historic_Fire_Perimeters_1516624541847049096.csv"), 
  colClasses = c('Cause' = 'factor',
                 'Collection.Method' = 'factor',
                 'Management.Objective' = 'factor',
                 'State' = 'factor',
                 'Agency' = 'factor',
                 'Unit.ID' = 'factor'))

#formatting date columns
CA_fires$Alarm.Date <- mdy_hms(CA_fires$Alarm.Date)
CA_fires$Containment.Date <- mdy_hms(CA_fires$Containment.Date)

CA_fires <- CA_fires %>%
  mutate(Cause = recode(Cause, 
                        "1" = "Lightning",
                        "2" = "Equipment Use",
                        "3" = "Smoking",
                        "4" = "Campfire",
                        "5" = "Debris",
                        "6" = "Railroad",
                        "7" = "Arson",
                        "8" = "Playing With Fire",
                        "9" = "Miscellaneous",
                        "10" = "Vehicle",
                        "11" = "Powerline",
                        "12" = "Firefighter Training",
                        "13" = "Non-Firefighter Training",
                        "14" = "Unidentified",
                        "15" = "Structure",
                        "16" = "Aircraft",
                        "17" = "Volcanic",
                        "18" = "Escaped Prescribed Burn",
                        "19" = "Illegal Alien Campfire"
                        ))


#reading in shapefile
fires_shapefile <- st_read("Data/raw/California_Fire_Perimeters_(1950+).shp")
CA_boundary_shp <- st_read("Data/spatial/CA_State.shp")

write.csv(CA_fires, file=here("./Data/processed/CA_fires.csv"))
```

# Exploratory Data Analysis

Our first step was to look at the data to look for any overall trends or notable patterns, to better inform our analysis. The following plots reveal how the fire sizes and frequency have occurred throughout the past 75 years, as well as what the top five fire causes are. 

```{r theme set, include = FALSE}
mytheme <- theme_bw(base_size = 12) +
  theme(axis.text = element_text(color = "black"),legend.position = "top", 
        plot.title = element_text(face = "bold"))
theme_set(mytheme)

```

First, this shows the column names and a few data points for our table we are working with.
```{r data summary, echo=FALSE}
#View dimensions, column names, variable type, and head of each column:
str(CA_fires)

head(CA_fires)
#this prints really weird - maybe we just do the column names
```

```{r histogram of fire counts, echo=FALSE, fig.cap= "Number of Fires Over Time"}
fire_count_plot <- CA_fires %>%
  ggplot(aes(x = Year)) + 
  labs( x = "Year",
        y = "Number of Fires",
        title = "Number of Fires Over Time") + 
  geom_histogram(binwidth = 1,
                 fill = 'firebrick3',
                 color = 'darkred') + 
  scale_x_continuous(n.breaks = 7)

print(fire_count_plot)
```

This plot shows the number of fires per year for the past 75 years, and there seems to be an upward trend in the number of fires, with some years having large spikes signaling and unusual amount of fires for that year.

```{r scatter plot of fire burn size, echo=FALSE, fig.cap= "Burn Size (Acres) by Year"}

options(scipen = 999) #this command keeps the y axis from switching into scientific notation

fire_size_plot <- CA_fires %>% 
  ggplot(aes(y=GIS.Calculated.Acres, x=Year)) + 
  geom_point(fill = 'firebrick3', color = 'darkred', alpha = 0.5) + 
  scale_y_continuous(labels = label_comma(),  limits = c(0,500000)) +
  labs(y="Burn Acreage (Calculated in GIS)", title = "Burn Size (Acres) by Year")

print(fire_size_plot)
```

Though the majority of fires appear to be less than 50,000 acres, we are interested in the scatter of dots in the 300,000 to 500,000 acre range that has increased since 1990. This could mean that burn acreage for fires in California is increasing over time.

```{r, echo=FALSE, fig.cap = "Total Fires Started by Each Cause, 1950 - 2025"}

fire_cause_scatter <- CA_fires %>% 
  ggplot(aes(y = Year, x = Cause)) + 
  geom_point(color = 'firebrick3', alpha = 0.1, size = 2.5) + 
  labs(y="Year", 
       title = "Total Fires Started by Each Cause, 1950 - 2025") +
  theme(axis.text.x = element_text(angle = 45, , vjust = 1, hjust = 1)) 
  
print(fire_cause_scatter)
```

This scatterplot is helpful because it shows wildfire events by ignition cause over time, with each point representing a single fire from 1950–2025. Denser vertical groupings indicate causes that occur more frequently, while sparser columns represent less common ignition sources. Causes that span the full time range appear persistent across decades (Debris and Equipment Use), whereas those concentrated in certain periods (Vehicles and Powerlines) suggest more episodic occurrence or changing patterns over time.

```{r bar graph of fire causes,echo=FALSE, fig.cap = "Total Fires Started by Each Cause, 1950 - 2025"}

fire_cause_count <- CA_fires %>% 
  ggplot(aes(x = fct_infreq(Cause))) +
  geom_bar(fill = "firebrick3", color = "darkred") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Cause",
       y = "Number of Fires",
       title = "Total Fires Started by Each Cause, 1950 - 2025")

print(fire_cause_count)
```

This plot reveals that the top five most commonly observed fire causes are:
  1. Unidentified 
  2. Lightning
  3. Miscellaneous
  4. Equipment Use
  5. Arson

The metadata does not describe what is meant by Unidentified or Miscellaneous, but we feel these categories are too large to ignore. These categories are likely due to the fact that much of this data is very old and there might not have been much work to identify the cause of past fires. 

```{r box plots of fire causes, echo=FALSE, fig.cap = "Top Five Fire Causes Over Time"}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Unidentified" | Cause == "Lightning" | Cause == "Miscellaneous" | Cause == "Equipment Use" | Cause == "Arson") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year, color = Cause)) + 
  geom_line(linewidth = 0.75) +
  geom_point() +
  scale_color_brewer(palette = 'Paired') +
  labs(y="Number of Fires", 
       x="Year",
       title = "Top Five Fire Causes Over Time")  + 
  theme(legend.position =  "right")

print(fire_cause_plot)
```

This plot notably reveals that there appears to be an overall increasing trend in the fires caused by lightning, which is the second leading cause of fires in California. In the following plot, we focused in on the trend in lighting-caused fires, and there is a clear increase. This informs our focus on fires caused by lightning for the remainder of the analysis.

```{r fires caused by lightning, echo=FALSE, fig.cap = "Number of Fires Caused by Lightning Strikes Over Time"}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year)) + 
  geom_line(linewidth = 0.75, color = 'darkred') +
  geom_point(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick2') +
  labs(y="Number of Fires", 
       x="Year",
       title = "Number of Fires Caused by Lightning Strikes Over Time")  + 
  theme(legend.position =  "right")

print(fire_cause_plot)
```

```{r mapview, echo=FALSE, fig.cap="Map of Fires 1950 - 2025"}
# visualize overlapping fire areas using symbology to show where fires are concentrated

fire_map <- ggplot() + 
  geom_sf(data = CA_boundary_shp) +
  geom_sf(data = fires_shapefile, fill = 'firebrick3', color = 'darkred', alpha = 0.5) +
  labs(title = "Map of Fires 1950 - 2025")

print(fire_map)
```

This map shows an overlay of all of the fire burn ares for the past 75 years. Areas in dark red show where multiple burns have occurred throughout the years. It is interesting to see the concentration of fires where California's mountain ranges are, and how the Central Valley is skipped by fire.

# Hypotheses
From this exploratory analysis, we developed the following hypotheses:
  **Alternative Hypothesis: The burn acreage of California wildfires has increased over time.**
    Null hypothesis: There has been no change in burn acreage of California wildfires over time.

  **Alternative Hypothesis: The number of California wildfires caused by lightning strikes has increased over time**
    Null hypothesis: There has been no change in the frequency of California wildfires caused by lightning over time.

# Analysis
Our analysis looked at how wildfire activity has changed over time, focusing on total burned area and fires caused by lightning. For burned area, monthly fire data were added up to get yearly totals. The data were then analyzed at both yearly and monthly scales to see if changes over time were overall trends or driven by seasonal patterns. A linear regression model (tslm) (or MK test??) was used for the yearly data because there was no clear seasonal pattern, allowing us to directly test whether burned area increased over time.

For the monthly burned area data, a time series decomposition was used to split the data into trend, seasonal patterns, and random variation. We then ran a Seasonal Mann–Kendall to test whether there was a real trend over time after accounting for seasonality.

The same approach was used for lightning-caused fires. Monthly fire counts were summed and analyzed over time. Yearly data were tested using a Mann–Kendall test since there was no seasonal pattern at that scale. Monthly data were again broken down using decomposition, and a Seasonal Mann–Kendall test was used to check for trends while accounting for seasonal changes.

## Question One - Burn Acreage Over Time

Our first analysis focuses on if fire acreage in California has increased, which would be due to an increase in overall fires or an increase in large fires. We first created a time series for the yearly burn acreage over time, the plotted it.
```{r burn acreage yearly time series plot, echo=FALSE, fig.cap="Time Series Plot: Yearly Burn Acreage Over Time"}
#adding up acres burned per year and checking that every year is in there
fire_yearly <- CA_fires %>% 
  group_by(Year) %>% 
  summarise(total_acres = sum(GIS.Calculated.Acres, na.rm = TRUE)) %>% 
  complete(Year = 1950:2025, fill = list(total_acres = 0))

# creating annual time series
fire_acres_ts <- ts(fire_yearly$total_acres,
                    start = 1950,
                    end = 2025,
                    frequency = 1)

fire_yearly_ts_plot <- fire_yearly %>%
  ggplot(aes(x = Year , y = total_acres)) +
  geom_line(color = 'darkred') +
  geom_smooth(method = "lm", color = 'firebrick2') +
  labs(y = "Yearly Burn Acreage",
       title = "Time Series Plot: Yearly Burn Acreage Over Time")
print(fire_yearly_ts_plot)
``` 
We then ran a linear regression to test for any linear trends in the yearly data.
**do we want to keep the tslm or just use the man kendall?**

```{r burn acreage yearly time series linear regression, echo=FALSE }
# can't decompose yearly sum because its not seasonal so running a linear regression
# chose tslm over lm because tslm accounts for trends and lm does not
fire_lm <- tslm(fire_acres_ts ~ trend)
summary(fire_lm)

#mann kendall
non_seasonal_acreage <- MannKendall(fire_acres_ts)
summary(non_seasonal_acreage)
```

Then, the fire acreage was calculated per month to create a time series analysis focused on seasonal trend analysis.

```{r burn acreage monthly time series, echo=FALSE, warning=FALSE, fig.cap="Time Series Plot: Monthly Burn Acreage Over Time"}
#now looking at monthly trend over time - seasonal so I can decompose it
CA_fires_2 <- CA_fires %>% 
  mutate(Alarm.Date = as.Date(Alarm.Date)) %>%
  mutate(Month = floor_date(Alarm.Date, unit = "month")) %>% #creates a monthly grouping
  group_by(Month) 

#creatly monthly sum
fire_monthly <- CA_fires_2 %>%
  summarise(total_acres = sum(GIS.Calculated.Acres, na.rm = TRUE)) %>%ungroup()

#create time series
monthly_acres_ts <- ts(fire_monthly$total_acres, start = c(1950, 1), frequency = 12)

fire_monthly_ts_plot <- fire_monthly %>%
  ggplot(aes(x = Month , y = total_acres)) +
  geom_line(color = 'darkred') +
  labs(y = "Monthly Burn Acreage",
       title = "Time Series Plot: Monthly Burn Acreage Over Time")
print(fire_monthly_ts_plot)
```

We ran a decomposition on this then a Seasonal Mann-Kendall test.

```{r burn acreage monthly time series2, echo=FALSE, fig.cap= "Time Series Decomposition: Monthly Burn Acreage Over Time"}
#decompose
monthly_acres_decomp <- stl(monthly_acres_ts, s.window = "periodic")
plot(monthly_acres_decomp,
      main = "Time Series Decomposition: Monthly Burn Acreage Over Time")

#seasonal = running seasonal mann-kendall test for trends
monthly_trend <- trend::smk.test(monthly_acres_ts)
monthly_trend
summary(monthly_trend)
```

This Seasonal Mann–Kendall test shows weak evidence of an increasing trend in monthly wildfire acreage over time (z = 1.648), but the trend is not statistically significant (p = 0.099). Seasonal decomposition doesn't show a strong trend within individual months, but some of the summer months have slightly positive but non-significant increases. For example, July (tau = 0.148). These results show variability and occasional increases in wildfire activity, but not a large, consistent seasonal trend across all months. This contrasts with the trend shown when looking at yearly fire acreage.

The yearly linear regression shows a clear, statistically significant increase in burned acreage over time, with a positive coefficient indicating rising fire extent and an R² of 0.1803, meaning about 18% of the variation in burned area is explained by time alone (p < 0.05). The monthly Seasonal Mann-Kendall test shows weaker evidence of a trend (z = 1.648, p = 0.099), suggesting that increases are not consistently strong at the monthly scale once seasonality is accounted for. The seasonal decomposition supports this: the trend component shows a gradual upward slope with occasional sharp jumps that likely reflect extreme fire years or major shifts in conditions, the seasonal component shows a stable, repeating annual fire cycle, and the remainder highlights large spikes in activity that represent unusually severe fire events. Overall, the results suggest that wildfire activity is increasing over time, but much of the change is driven by long-term shifts and extreme years rather than a steady month-to-month increase.

## Question Two - Lightning Frequency Over Time

Informed by the exploratory analysis, we analyzed if the amount of fires caused by lightning strikes has changed over time. This can help determine if fire trends are related to increasing lightning storms as a result of climate change.

Similar to the first research question, the total count of fires caused by lighting was first summed by year to create a time series analysis before running a Mann Kendall test.
```{r lightning yearly time series, echo=FALSE}

lightning_fire_yearly <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  mutate(Year = as.factor(Year)) %>%
  group_by(Year) %>% 
  mutate(Year = as.Date(Year, format = "%Y"))  %>%
  summarise(count = n()) 

#creating annual time series
lightning_fire_yearly_ts <- ts(lightning_fire_yearly,
                    start = 1950,
                    end = 2025,
                    frequency = 1)

#removing time series plot because it matches the one from our exploratory analysis

#mann kendall
non_seasonal_lightning <- MannKendall(lightning_fire_yearly_ts)
summary(non_seasonal_lightning)
```

A Mann-Kendall trend test found a statistically significant increasing trend in the annual number of lightning-caused fires over time (tau = 0.111, p = 0.043). This indicates that lightning-caused fire counts generally increased across the study period, although the strength of the trend was relatively weak.

The following analysis looks at the sum of fires caused by lightning per month, to look at an seasonal change. For this time series, we ran a decomposition and a Seasonal Mann Kendall test.

```{r lightning monthly time series2, echo=FALSE, fig.cap = "Time Series Plot: Monthly Fires Caused by Lightning Over Time" }
#now looking at monthly trend over time - seasonal so I can decompose it
lightning_fire_monthly <- CA_fires %>% 
  filter (Cause == "Lightning") %>% 
  mutate(Alarm.Date = as.Date(Alarm.Date)) %>%
  mutate(Month = floor_date(Alarm.Date, unit = "month")) %>% #creates a monthly grouping
  group_by(Month) 

#create monthly sum
lightning_fire_monthly <- lightning_fire_monthly %>%
  summarise(count = n())

#creating data frame
Month <- as.data.frame(seq(from = as.Date("1950-05-01"),
                           to = as.Date("2025-01-01"), by = "month"))

#renaming column
Month <- rename(Month, 'Month' = 'seq(from = as.Date(\"1950-05-01\"), to = as.Date(\"2025-01-01\"), by = \"month\")')

#joining
lightning_fire_monthly_join <- left_join(Month, lightning_fire_monthly, by = "Month")

#linear interpolation
lightning_fire_monthly_join$count <- na.approx(lightning_fire_monthly_join$count, rule = 2)

#create time series
lightning_monthly_ts <- ts(lightning_fire_monthly_join$count, start = c(1950, 1), frequency = 12)
plot(lightning_monthly_ts)


lightning_monthly_ts_plot <- lightning_fire_monthly_join %>%
  ggplot(aes(x = Month , y = count)) +
  geom_line(color = 'darkred') +
  labs(y = "Monthly Fire Count",
       title = "Time Series Plot: Monthly Fires Caused by Lightning Over Time")
print(lightning_monthly_ts_plot)
```


```{r lightning monthly time series3, echo=FALSE, fig.cap ="Time Series Decomposition: Monthly Count of Fires Caused by Lightning" }
#decompose
lightning_monthly_decomp <- stl(lightning_monthly_ts, s.window = "periodic")
plot(lightning_monthly_decomp,
     main = "Time Series Decomposition: Monthly Count of Fires Caused by Lightning")


#seasonal = running seasonal mann-kendall test for trends
lightning_monthly_trend <- trend::smk.test(lightning_monthly_ts)
lightning_monthly_trend
summary(lightning_monthly_trend)
```

The decomposition shows that lightning-caused fires follow a regular yearly pattern, with more fires in some months and fewer in others. This means the timing of these fires is fairly consistent from year to year. It also shows that the overall number of fires has gradually increased over time, with more recent years generally having higher counts. The leftover ups and downs represent unusual months with especially high or low fire activity that are not part of the normal seasonal pattern.

A Seasonal Mann-Kendall test showed that the monthly number of lightning-caused fires has increased over time, even when normal seasonal patterns are taken into account (z = 7.42, p < 0.001). The increase was strongest during late winter and spring, especially in Seasons 2 through 5 (February - May), with Season 3 (March) showing the largest rise. Most other months also showed upward trends, but those increases were smaller and not statistically significant.

# Summary 
summary for yearly fire acreage

summary for monthly fire acreage 

summary for yearly lightning

summary for monthly lightning

# Conclusions
what this means for fires in CA and our focus on cliamte change 
maybe a policy reccomendation or something??

# Limitations
It is noted in this data's metadata that there are a number of fires missing from these records, mostly for the first few decades of the dataset. With this, the trend analysis might be skewed due to the missing data. This is important to note when drawing conclusions from the data, especially when focusing the on yearly and monthly fire acreage. 

The two categories of "Unidentified" and "Miscellaneous" are also a limitation, and it would be interesting to run the exploratory analysis again if the fires in these categories are correctly grouped by specific causes. 

## References
Fire Data:
URL: https://gis.data.cnra.ca.gov/datasets/CALFIRE-Forestry::california-fire-perimeters-1950/about
Citation Title: Fire Perimeters (firep24_1) 
Revision date: 2025-04-23 
Edition: 24_1 
Processing environment Version 6.2 (Build 9200); Esri ArcGIS 10.8.0.12790 

CA Boundary Shapefile:
URL: https://data.ca.gov/dataset/e212e397-1277-4df3-8c22-40721b095f33/resource/3db1e426-fb51-44f5-82d5-a54d7c6e188b/download/ca_state.zip
California State boundary in shapefile format from the US Census Bureau's 2023 MAF/TIGER database.