---
title: "ENV 782 Final - CA Wildfires"
author: "Sky Volz and Brianna Karson"
date: "2026-04-19"
output: pdf_document
---

\tableofcontents
\listoffigures

# Rationale and Introductions

Wildfires are a pressing environmental challenge in California that has been exacerbated by decades of fire suppression, increasing drought conditions, and a changing climate. This dataset contains wildfire data for the past 75 years, offering a unique glimpse into how fire patterns have shifted over time. This report focuses on how climate change has impacted wildfires, with increasingly intense droughts contributing to larger fires, or with increased frequency of storms potentially causing more fires via lightning striking dry fire fuels. By providing statistical evidence that wildfire or lightning strike intensity has increased over time, this research can inform policy recommendations, ideally encouraging more proactive fire management strategies such as prescribed burning to reduce fuels.

# Research Questions

This project explores the following research questions: 

  **1. Has the size of California wildfire burn areas of changed over time?**
  
  **2. Has the frequency of California wildfires caused by lightning strikes changed over time?**

# Dataset Information

This dataset is sourced from the California Natural Resources Agency GIS database and is serviced by the California Department of Forestry and Fire Protection’s  CAL FIRE eGIS. It includes layers that subset from the larger dataset according to year the fire occurred, and size of the burn area in acres. It contains 17,540 records, spanning from the 1950s to May 9, 2025. It is a publicly available feature layer and was last updated on May 9, 2025.

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
#loading packages 
library(tidyverse)
library(dplyr)
library(here)
library(sf)
library(ggplot2)

#set working directory
here()
```

```{r data wrangling, message=FALSE, warning=FALSE, include=FALSE}
#reading in data
CA_fires <- read.csv(file=here(
  "Data/raw/California_Historic_Fire_Perimeters_1516624541847049096.csv"), 
  colClasses = c('Cause' = 'factor',
                 'Collection.Method' = 'factor',
                 'Management.Objective' = 'factor',
                 'State' = 'factor',
                 'Agency' = 'factor',
                 'Unit.ID' = 'factor'))

#formatting date columns
CA_fires$Alarm.Date <- mdy_hms(CA_fires$Alarm.Date)
CA_fires$Containment.Date <- mdy_hms(CA_fires$Containment.Date)

CA_fires <- CA_fires %>%
  mutate(Cause = recode(Cause, 
                        "1" = "Lightning",
                        "2" = "Equipment Use",
                        "3" = "Smoking",
                        "4" = "Campfire",
                        "5" = "Debris",
                        "6" = "Railroad",
                        "7" = "Arson",
                        "8" = "Playing With Fire",
                        "9" = "Miscellaneous",
                        "10" = "Vehicle",
                        "11" = "Powerline",
                        "12" = "Firefighter Training",
                        "13" = "Non-Firefighter Training",
                        "14" = "Unidentified",
                        "15" = "Structure",
                        "16" = "Aircraft",
                        "17" = "Volcanic",
                        "18" = "Escaped Prescribed Burn",
                        "19" = "Illegal Alien Campfire"
                        ))


#reading in shapefile
fires_shapefile <- st_read("Data/raw/California_Fire_Perimeters_(1950+).shp")

#save this processed data table to processed folder - idk why this wont work
#write.csv(CA_fires, file=here("/home/Workspace/Gits/SkyVolz_BriannaKarson/Data/processed"))
```

# Exploratory Data Analysis

Our first step was to look at the data to look for any overall trends or notable patterns, to better inform our analysis. The following plots reveal how the fire sizes and frequency have occurred throughout the past 75 years, as well as what the top five fire causes are. 

```{r theme set}
mytheme <- theme_minimal(base_size = 12) +
  theme(axis.text = element_text(color = "black"),legend.position = "top", 
        plot.title = element_text(face = "bold"))
theme_set(mytheme)

#should we hide this code?
```

```{r}
#View dimensions, column names, variable type, and head of each column:
str(CA_fires)

summary(CA_fires$GIS.Calculated.Acres)
```

```{r histogram of fire counts, fig.cap = "Changes in Fire Occurences Over Time"}
fire_count_plot <- CA_fires %>%
  ggplot(aes(x = Year)) + 
  labs( x = "Year",
        y = "Number of Fires",
        title = "Changes in Fire Occurences Over Time") + 
  geom_histogram(binwidth = 1,
                 fill = 'firebrick3',
                 color = 'darkred') + 
  scale_x_continuous(n.breaks = 7)

print(fire_count_plot)
```

```{r scatter plot of fire burn size, fig.cap = "Burn Size (Acres) by Year"}
fire_size_plot <- CA_fires %>% 
  ggplot(aes(y=GIS.Calculated.Acres, x=Year)) + 
  geom_point(fill = 'firebrick3', color = 'darkred', alpha = 0.5) + 
  #idk was just trying stuff with the colors
  labs(y="Burn Acreage Calculated in GIS", title = "Burn Size (Acres) by Year") + 
  ylim(0,500000)  

#i agree no outliers and no trend line!
#what if we filter out to show fires 500,000+ to show just big fires?

print(fire_size_plot)
```
```{r bar graph of fire causes, fig.cap = "Total Fires Started by Each Cause, 1950 - 2025"}
fire_cause_count <- CA_fires %>% 
  ggplot(aes(x=Cause)) +
  geom_bar(fill = 'firebrick3', color = 'darkred') + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(y = "Count", title = "Total Fires Started by Each Cause, 1950 - 2025") +
  geom_text(stat = 'count', 
            aes(label = after_stat(count)), 
            vjust = -0.5, size = 2)

print(fire_cause_count)

#i think we explain that this shows the overall count, then use it to show the top five in the next plot -  but should we exclude miscellanous and unidentified?

```

This plot reveals that the top five fire causes according to frequency are:
  1. Unidentified 
  2. Lightning
  3. Miscellaneous
  4. Equipment Use
  5. Arson

```{r box plots of fire causes, fig.cap = "Changes in Frequency of the Top Five Fire Causes"}
fire_cause_plot <- CA_fires %>% 
  filter (Cause == "Unidenfitied" | Cause == "Lightning" | Cause == "Miscellaneous" | Cause == "Equipment Use" | Cause == "Arson") %>% 
  count(Year, Cause) %>%  
  ggplot(aes(y=n, x=Year, color = Cause)) + 
  geom_line(linewidth = 1) +
  geom_point() +
  scale_color_brewer(palette = 'RdYlBu') +
  labs(y="Cause of Fire", 
       x="Year",
       title = "Changes in Frequency of the Top Five Fire Causes") 

print(fire_cause_plot)

#I like this because plot it shows that lightning is a big issue and is increasing so it is a good reason to focus on it for the rest of the report
#change these colors?
```

This plot notably reveals that there appears to be an overall increasing trend in the fires caused by lightning, which is the second leading cause of fires in California. This informs our focus on fires caused by lightning for the remainder of the analysis.

```{r mapview, fig.cap="Map of Fires 1950 - 2025"}
# visualize overlapping fire areas using symbology to show where fires are concentrated
fire_map <- ggplot(data = fires_shapefile) + 
  geom_sf( fill = 'firebrick3', color = 'darkred', alpha = 0.5)

print(fire_map)

```

# Hypotheses
From this exploratory analysis, we developed the following hypotheses:
  **1. The burn acreage of wildfires has increased over time.**
  (add null and alt hyp here?)
  **2. The frequency of fires caused by lightning strikes has increased over time**
   (add null and alt hyp here?)

# Analysis
We will state our methods here

## Question One - Burn Acreage Over Time
```{r burn acreage time series}

#Time Series Analysis: How has the burn acreage of wildfires changed over time?
  #Area as the dependent variable
  #Decomposition and trend analysis
```


## Question Two - Lightning Frequency Over Time
```{r lightning time series}

#Time Series Analysis: Has the frequency of fires caused by lightning strikes changed over time?
  #Dependent variable is the count of lightning strikes
  #Decomposition and trend analysis
```

# Summary and Conclusions
# Results
# Discussion
# Limitations
see “Use Limitation” section in metadata

## References
Remember to cite dataset here 